---
title: "P8106_midterm"
author: "yimin chen yc4195"
date: "2023-03-26"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = T, message = FALSE, results='hide', warning=FALSE}
library(corrplot)
library(tidyverse)
library(caret)
library(mgcv)
library(patchwork)
library(rpart)
library(rpart.plot)
library(party)
library(randomForest)
library(ranger)
library(e1071)
library(pdp)
library(earth)
library(splines)
library(pdp)
library(ggplot2)
library(gridExtra)
library(glmnet)
library(MASS)
library(pROC)
library(vip)
library(AppliedPredictiveModeling)
library(klaR)
```


## import and Data
```{r, import}
set.seed(4195) 
load("recovery.Rdata")
dat <- dat[sample(1:10000, 2000),]
dat <- dat[,-1]%>% #ignore the id variable
    mutate(study = case_when( # from character variable to a numeric variable
    study == "A" ~ 1,
    study == "B" ~ 2,
    study == "C" ~ 3)) %>%
    mutate(
    gender = as.factor(gender),
    race = as.factor(race),
    smoking = as.factor(smoking),
    hypertension = as.factor(hypertension),
    diabetes = as.factor(diabetes),
    vaccine = as.factor(vaccine),
    severity = as.factor(severity),
    study = as.factor(study)
  )
skimr::skim_without_charts(dat)
```

## Create x and y matrixs for modeling
```{r, model_matrices}
ctrl <- trainControl(method = "repeatedcv",
                     repeats = 5,
                     number=10)
set.seed(4195)
traindata <- createDataPartition(dat$recovery_time, p = 0.7, list = FALSE)
traindataset <- dat[traindata,]
testdataset <- dat[-traindata,]
#train
x = model.matrix(recovery_time ~ ., dat)[traindata,-1]
y = traindataset$recovery_time

#test
x_test = model.matrix(recovery_time ~ ., dat)[-traindata,-1]
y_test = testdataset$recovery_time

```


# Exploratory analysis and data visualization:

```{r}
#corr plot
corrplot(cor(x), method = "circle", type = "full")

# create dataset for exploratory analysis and data visualization
traindataset1 <- traindataset%>%
    mutate(
    gender = as.numeric(gender),
    race = as.numeric(race),
    smoking = as.numeric(smoking),
    hypertension = as.numeric(hypertension),
    diabetes = as.numeric(diabetes),
    vaccine = as.numeric(vaccine),
    severity = as.numeric(severity),
    study = as.numeric(study))

```

```{r,warning=FALSE}
theme1 = trellis.par.get()
theme1$plot.symbol$col = rgb(.2, .4, .2, .5)
theme1$plot.symbol$pch = 16
theme1$plot.line$col = rgb(.8, .1, .1, 1)
theme1$plot.line$lwd = 2
theme1$strip.background$col = rgb(.0, .2, .6, .2)
trellis.par.set(theme1)

## Exploratory analysis and visualization
featurePlot(x = traindataset1[ ,1:14],
            y = traindataset1[ ,15],
            plot = "scatter",
            span = .5,
            #layout=c(4,4),
            labels = c("Predictors", "Recovery Time"),
            type = c("p", "smooth"))
```


# Model training:


## Fit a linear model:

```{r lm,cache=TRUE}
set.seed(4195)
# Fit a linear regression model 
lm.fit <- train(recovery_time ~ age + gender + race + smoking + height + weight + 
                 bmi + hypertension + diabetes + SBP + LDL + vaccine + 
                 severity + study, 
               data = traindataset, 
               method = "lm", 
               trControl = ctrl)
# model summary
summary(lm.fit$finalModel)
# RMSE
test_pred_lm <- predict(lm.fit, newdata = testdataset)
test_rmse_lm <- sqrt(mean((test_pred_lm -y_test)^2))
test_rmse_lm
```
RMSE = `r test_rmse_lm `

                 
## Fit Ridge Regression
```{r, ridge, cache=TRUE}
set.seed(4195)
ridge.fit = train(x, y,
                  method = "glmnet",
                  tuneGrid = expand.grid(alpha = 0,
                                         lambda = exp(seq(-10, 6, length = 200))),
                  trControl = ctrl)
ridge_pred = predict(ridge.fit$finalModel, newx = x_test, s = ridge.fit$bestTune$lambda, type = "response")
# model summary
summary(ridge.fit$finalModel)
plot(ridge.fit,xTrans =log)
ridge.fit$bestTune
# RMSE
test_pred_ridge <- predict(ridge.fit, newdata = data.frame(x_test))
test_rmse_ridge <- sqrt(mean((test_pred_ridge -y_test)^2))
test_rmse_ridge
```
RMSE = `r test_rmse_ridge `


## Fit Lasso model:

```{r, lasso, cache=TRUE}
set.seed(4195)
lasso.fit <- train(x, y,
                   method = "glmnet",
                   tuneGrid = expand.grid(alpha = 1,
                                          lambda = exp(seq(-5, -2, length = 100))),
                   trControl = ctrl)
# view the model summary
summary(lasso.fit$finalModel)
plot(lasso.fit,xTrans =log)
lasso.fit$bestTune
# view performance on the test set (RMSE)
lasso_pred <- predict(lasso.fit, newdata = data.frame(x_test))
test_rmse_lasso<- sqrt(mean((lasso_pred - y_test)^2))
test_rmse_lasso
```
RMSE = `r test_rmse_lasso `

## Fit Elastic net model:

```{r,enet,cache=TRUE}
set.seed(4195)
enet.fit <- train(x, y,
                  method = "glmnet",
                  tuneGrid = expand.grid(alpha = seq(0, 1, length = 21), 
                                         lambda = exp(seq(2, -2, length = 50))),
                  trControl = ctrl)
# view the model summary
summary(enet.fit$finalModel)
enet.fit$bestTune
myCol<- rainbow(25)
myPar <- list(superpose.symbol = list(col = myCol),
                    superpose.line = list(col = myCol))

plot(enet.fit, par.settings = myPar)

coef(enet.fit$finalModel, enet.fit$bestTune$lambda)
# view performance on the test set (RMSE)
enet_pred <- predict(enet.fit, data.frame(x_test))
test_rmse_enet <- sqrt(mean((enet_pred - y_test)^2))
test_rmse_enet
# calculate RMSE
sqrt(mean((enet_pred - y_test)^2)) 
```
RMSE = `r test_rmse_enet `

## Fit a PCR model
```{r, pcr, cache=TRUE}
set.seed(4195)
pcr.fit <- train(x, y,
                 method = "pcr",
                 tuneGrid  = data.frame(ncomp = 1:19),
                 trControl = ctrl,
                 preProcess = c("center", "scale"))
# model summary
summary(pcr.fit$finalModel)
ggplot(pcr.fit, highlight = TRUE) + theme_bw()
# RMSE
test_pred_pcr <- predict(pcr.fit, newdata = data.frame(x_test))
test_rmse_pcr <- sqrt(mean((test_pred_pcr -y_test)^2))
test_rmse_pcr
```                 
RMSE = `r test_rmse_pcr ` 

## Fit a Partial least squares model(PLS):
```{r,pls,cache=TRUE}
set.seed(4195)
pls.fit <- train(x, y,
                 method = "pls",
                 tuneGrid = data.frame(ncomp = 1:15), 
                 trControl = ctrl,
                 preProcess = c("center", "scale"))
# view the model summary
summary(pls.fit$finalModel)
ggplot(pls.fit, highlight = TRUE) + theme_bw()
# view performance on the test set (RMSE)
pls_pred <- predict(pls.fit, data.frame(x_test))
test_rmse_pls <- sqrt(mean((pls_pred - y_test))^2)
test_rmse_pls
```
RMSE = `r test_rmse_pls `



## Fit a Generalized additive model (GAM):

```{r,gam,warning=FALSE,cache=TRUE}
set.seed(4195)
# fit GAM using default setting
gam.fit <- train(x, y, # test dataset
                 method = "gam",
                 trControl = ctrl)

# view the model summary
summary(gam.fit$finalModel)
plot(gam.fit)
gam.fit$bestTune
gam.fit$finalModel
par(mfrow=c(2, 3))
plot(gam.fit$finalModel)
par(mfrow=c(1, 1))
# view performance on the test set (RMSE)
gam_pred <- predict(gam.fit, data.frame(x_test))
test_rmse_gam <- sqrt(mean((gam_pred - y_test))^2)
test_rmse_gam


```

RMSE for gaml= `r test_rmse_gam `



## Fit a Multivariate adaptive regression spline (MARS) model:

```{r,mars,cache=TRUE}
set.seed(4195)
# create grid of all possible pairs that can take degree and nprune values
mars_grid <- expand.grid(degree = 1:3, # number of possible product hinge functions in 1 term
                         nprune = 2:20) # Upper bound of number of terms in model
mars.fit <- train(x,
                  y,# training dataset
                  method = "earth",
                  tuneGrid = mars_grid,
                  trControl = ctrl)
# view the model summary
summary(mars.fit$finalModel)
plot(mars.fit)
varImp(mars.fit)
# view performance on the test set (RMSE)
mars_pred <- predict(mars.fit, data.frame(x_test))
test_rmse_mars <- sqrt(mean((mars_pred - y_test))^2)
test_rmse_mars
```
RMSE = `r test_rmse_mars `



## Model comparison:

# Boxplot
```{r, boxplot}
set.seed(4195)
resamp = resamples(list(
                        linear=lm.fit,
                        ridge = ridge.fit,
                        lasso = lasso.fit,
                        enet = enet.fit,
                        pcr = pcr.fit,
                        pls = pls.fit,
                        gam = gam.fit,
                        mars = mars.fit))
summary(resamp)
bwplot(resamp, metric = "RMSE")

```


# choose GAM model



